The Hydrogen Epoch of Reionization Array Dish: Characterization with Electromagnetic Simulations

### 2016-01-13 DishSimulations/paper/DishSimulationsPaper_v1_old.tex

The foreground filtering procedure employed by PAPER and HERA involves delay transforming the visibilities on a short time-scale cadence and performing a 1d delay CLEAN where the deconvolution is permitted to discover and subtract foregrounds within the horizon to horizon delay. Since we want to avoid CLEANing noise, the level of foreground subtraction possible by this procedure is limited by the thermal noise level on the visibility, which in terun depends on the number of time steps and redundant baselines that are averaged before performing the cleaning step. In this work, we assume that each visibility is cleaned independently with a two minute CLEANing cadence which corresponds approximately to the time over which HERA's short baselines are coherent. The standard deviation on the real and imaginary part of a single delay transformed visibility is given by \citep{Morales:2004}
\begin{equation}
\Delta V = \frac{\sqrt{2 B} k_B T_{sys}}{A_e \sqrt{\tau}}
\end{equation}
where $A_e$ is the effective area of the dish, $B$ is the bandwidth, $T_{sys}$ is the system temperature, $\tau$ is the integration time, and $k_B$ is the Boltzmann constant. The system temperature can be calculated by assuming that $T_{sys} = 100\text{K} + T_{sky}$ where $100$\,K is the temperature of the PAPER reciever and $T_{sky} = 60 (\lambda/1\,\text{meter} )^{2.55}$ is the sky temperature \citep{Fixsen:2008}. For $A_e$ we use the value of $75$\,m determined in \citep{Neben:2015b}. If we assume that each baseline is cleaned independently and that the integration time is $\tau \approx 60$\,s, than the noise level at 150\,MHz is approximately $9.9$\,Jy\,MHz. In order to avoid subtracting noise, we assume that cleaning is performed down to $5\,\sigma$. In Fig~\ref{fig:Cleaning} we compare the delay transform of visibilities before and after cleaning on a two minute cadence at the LST of 4 hours. While cleaning is able to remove structure within the horizon, it does not reduce any of the power associated with the foreground induced by the chromaticity of the dish.

To form estimates of the 21\,cm power spectrum, we split each visibility into BlackmanHarris windowed sub-bands centered at redshift intervals of $\Delta z = 0.5$ and each with an equivalent bandwidth of $10\,MHz$, corresponding to the redshift interval over which the statistics of the brightness temperature fluctuations are expected to be stationary. For each windowed interval, we use the flat sky approximation and  Fourier transform in frequency, square, and multiply by a set of prefactors to obtain a power spectrum estimate \citep{Parsons:2014},
\begin{equation}\label{eq:PS}
\widehat{P}({\bf k}) = \left( \frac{2 k_B}{\lambda^2} \right)^2 \frac{X^2 Y}{B \Omega_{pp}} | \widetilde{V}({\bf u}) |^2.
\end{equation}
Here, $\lambda$ is the central wavelength of the observation, $k_B$ is the Boltzmann constant, $B$ is the bandwidth of the Fourier transform, and $\Omega_{pp}$ is the integrated solid angle of the primary beam squared. $X$ is a linear factor converting between $uv$ wavelengths and $k_\perp$, and $Y$ is a linear factor converting between $k_\parallel$ and the Fourier dual to frequency, $\eta$.

A drift scan instrument, HERA will be capable of observing the sky at any LST within the strip declination that passes through its primary beam. To form a final power spectrum estimate, it is expected to perform an average over the power spectrum estimates it obtains through equation~\ref{eq:PS} for each LST independently. It is well documented that foreground power is maximized over certain LST ranges \citep{Thyagarajan:2015a}, hence such an estimate will either filter or weight appropriately the LSTs in which foreground power is maximized. \citet{Thyagarajan:2015c} focuses on the optimal weighting and selection of LST bins in forming a power spectrum estimate. For the purpsoes of our analysis, we focus on a single, relatively clean LST of 4 hours which includes the coldest part of the sky.

@@

Computing the power spectrum, we inspect

\begin{figure*}
\includegraphics[width=\textwidth]{figures/ps_compare_nithya_lst4hr_1024ch_7p3ew_12p6ns.pdf}
\caption{The power spectrum computed from the delay transform of a 14.6 meter baseline in equation~\ref{eq:PS} with (solid green line) and without (solid blue line) the presence of reflections in the HERA dish. The residuals after cleaning within the dashed black lines down to three times the level of thermal noise after 20 minutes of integration with (dashed cyan) and without (red dashed) the dish chromaticity are also plotted. We find that for this short baseline, clean residuals are on a similar level as the 21\,cm signal (solid black line), indicating that a redesign of the array with fewer short baselines may be necessary to maximize its sensitivity on foreground free modes.}
\label{figure:shortBaseline}
\end{figure*}

\begin{figure*}
\includegraphics[width=\textwidth]{figures/ps_compare_nithya_lst4hr_1024ch_14p6ew_25p3ns.pdf}
\caption{The same as Fig.~\ref{fig:shortBaseline} except now for a 29.2 meter baseline. We see that the level of foregrounds is down relative to our 14.6 meter baseline by two orders of magnitude, likely due to the fact that we are now outside of the region with significant contamination from the Galaxy. The presence of reflections tends to extend the $k_\parallel$ at which foregrounds drop below the level of the signal between $k_\parallel=0.15$\,$h$Mpc$^{-1}$ and $k_\parallel=0.22$\,$h$Mpc$^{-1}$ with the most severe contamination occurring at the $z=8.5$ subband where the reflections where observed to be worst in Fig.~\ref{fig:KernelsSubbands}. }
\label{figure:shortBaseline}
\end{figure*}

>>>>

The foreground filtering procedure employed by PAPER and HERA involves delay transforming the visibilities on a short time-scale cadence and performing a 1d delay CLEAN where the deconvolution is permitted to discover and subtract foregrounds within the horizon to horizon delay. Since we want to avoid CLEANing noise, the level of foreground subtraction possible by this procedure is limited by the thermal noise level on the visibility, which in terun depends on the number of time steps and redundant baselines that are averaged before performing the cleaning step. In this work, we assume that each visibility is cleaned independently with a two minute CLEANing cadence which corresponds approximately to the time over which HERA's short baselines are coherent. The standard deviation on the real and imaginary part of a single delay transformed visibility is given by \citep{Morales:2004}
\begin{equation}
\Delta V = \frac{\sqrt{2 B} k_B T_{sys}}{A_e \sqrt{\tau}}
\end{equation}
where $A_e$ is the effective area of the dish, $B$ is the bandwidth, $T_{sys}$ is the system temperature, $\tau$ is the integration time, and $k_B$ is the Boltzmann constant. The system temperature can be calculated by assuming that $T_{sys} = 100\text{K} + T_{sky}$ where $100$\,K is the temperature of the PAPER reciever and $T_{sky} = 60 (\lambda/1\,\text{meter} )^{2.55}$ is the sky temperature \citep{Fixsen:2008}. For $A_e$ we use the value of $75$\,m determined in \citep{Neben:2015b}. If we assume that each baseline is cleaned independently and that the integration time is $\tau \approx 60$\,s, than the noise level at 150\,MHz is approximately $9.9$\,Jy\,MHz. In order to avoid subtracting noise, we assume that cleaning is performed down to $5\,\sigma$. The CLEAN is applied  independently to each LST. In Fig~\ref{fig:Cleaning} we compare the delay transform of visibilities before and after cleaning on a two minute cadence. While cleaning is able to remove structure within the horizon, it does not reduce any of the power associated with the foreground induced by the chromaticity of the dish.

To form estimates of the 21\,cm power spectrum, we split each visibility into BlackmanHarris windowed subbands centered at redshift intervals of $\Delta z = 0.5$ and each with an equivalent bandwidth of $\Delta z = 0.5$, corresponding to the redshift interval over which the statistics of the brightness temperature fluctuations are expected to be stationary. For each windowed interval, we use the flat sky approximation and  Fourier transform in frequency, square, and multiply by a set of prefactors to obtain a power spectrum estimate \citep{Parsons:2014},
\begin{equation}\label{eq:PS}
\widehat{P}({\bf k}) = \left( \frac{2 k_B}{\lambda^2} \right)^2 \frac{X^2 Y}{B \Omega_{pp}} | \widetilde{V}({\bf u}) |^2.
\end{equation}
Here, $\lambda$ is the central wavelength of the observation, $k_B$ is the Boltzmann constant, $B$ is the bandwidth of the Fourier transform, and $\Omega_{pp}$ is the integrated solid angle of the primary beam squared. $X$ is a linear factor converting between $uv$ wavelengths and $k_\perp$, and $Y$ is a linear factor converting between $k_\parallel$ and the Fourier dual to frequency, $\eta$.

A drift scan instrument, HERA will be capable of observing the sky at any LST within the strip declination that passes through its primary beam. To form a final power spectrum estimate, it is expected to perform an average over the power spectrum estimates it obtains through equation~\ref{eq:PS} for each LST independently. It is well documented that foreground power is maximized over certain LST ranges \citep{Thyagarajan:2015a}, hence such an estimate will either filter or weight appropriately the LSTs in which foreground power is maximized. To determine the LSTs to be included in our power spectra, we inspect the ratio between the power spectra foreground residuals after two minutes of cleaning and a model for the power spectrum of the 21\,cm signal for several baselines as a function of LST. We plot the ratio for one such baseline (at 14.6 meters, one from which the majority of HERA's sensitivity will be derived) in Fig.~\ref{fig:LSTs} and see that foreground power is at its worst during the transits of the galactic centers and anti-centers between LST of 8 and 20. In principle, one could form an average that weights inversely by foreground amplitudes allowing for very significant reduction in the overall foreground power, however such an approach also requires precise {\it a-priori} knowledge of the foregrounds which we will not assume here. Instead, we take a conservative approach and form LST averaged power spectra by taking an unweighted average over all LSTs excluding the highly contaminated region between 8 and 20 hours.

@@

In Fig.~\ref{fig:SignalCompare} we compare the levels of the LST averaged foregrounds and cleaning residuals with and without the chromaticity of the dish for the redshifts between $7$ and $10.5$. Between the redshifts of $7$ and $8.5$, the residuals with and without the chromaticity of the dish drop below the level of the 21\,cm signal below $k \approx 0.12-0.15$\,$h$Mpc$^{-1}$. At $z = 8.5$, at the band center, where we found the reflections to be most severe, the beam chromaticity extends foreground contamination in excess of the signal out to $k \approx 0.22$\,$h$Mpc$^{-1}$. As we move to progressively higher redshift, the fixed frequency interval corresponding to $\Delta z = 0.5$ decreases as $(1+z)^{-2}$ while the foregrounds themselves also become brighter. The power spectrum we use here also has also reached its maximum $z=8.5$ where $50\%$ reionization takes place. For these reasons, the $k_\parallel$ at which the foregrounds with and without beam chromaticity fall below the signal increase with redshift, falling below the between $k=0.17$ and $k=0.22$\,$h$Mpc$^{-1}$ out to $z=10.5$.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/snr_estimates_subBands_Nithya_comovingDF.png}
\caption{The ratio between the foreground power spectrum and the power spectrum of 21\,cm brightness temperature fluctuations on a single $14.6$\,m baseline for several different redshift subbands assuming two minutes of foreground cleaning and chromaticity of our simulated beam. Significant foreground contamination enters between 8 and 20 hours, hence we form LST averaged power spectra excluding this time interval (denoted by vertical dashed red lines).}
\label{fig:LSTs}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/PS_estimates_subBands_Nithya_comovingDF_avgMask.pdf}
\caption{The delay transformed power spectra of foregrounds with (blue line) and without (green line) contamination from reflections in the dish. The residuals of these foregrounds, cleaned to $5\,\sigma$ after 2 minutes of integration, are plotted with (red dashed line) and without (cyan dashed line) dish reflections. The power spectra are derived from averaging over all LST excluding 8-20 hours. We compare to the amplitude of the 21\,cm power spectrum (solid black line). For each redshift, we use an equivalent bandwidth (for a BlackmanHarris window) of the frequency range corresponding to $\Delta z = 0.5$. We find that over most subbands, the foreground leakage extended by the chromaticity of the dish is below $0.12$-$0.15$\,$h$Mpc$^{-1}$. However, in the center of the band, where the reflections are significantly pronounced, the foreground leakage does not pass below $\sim 10$\% the level of the signal until  $k_\parallel \approx 0.3$\,$h$Mpc$^{-1}$.}
\label{fig:SignalCompare}
\end{figure*}
